---
title: "DALEX docs archived objects"
author: "Alicja Gosiewska"
date: "17 May 2018"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=6, results='asis', warning=FALSE, message=FALSE, cache=FALSE)
```



```{r}
library(archivist)
library(DALEX)
library(gbm)
library(ggplot2)
library(knitr)
library(randomForest)
```


```{r, echo = FALSE}
addHook <- function(object){
  name <- deparse(substitute(object))
  hook <- asave(object, repoDir = "DALEX_arepo", artifactName = name)
cat("Load ", name,": [`archivist::aread('pbiecek/DALEX_arepo/", hook, "')`](https://raw.githubusercontent.com/pbiecek/DALEX_arepo/master/gallery/", hook, ".rda)"," \n \n", sep = "")
}
```

# Chapter 2 Architecture of DALEX

## 2.2 Use case: Regression. Apartment prices in Warsaw

```{r, results = 'hide', fig.show = FALSE}
library(DALEX)
head(apartments)
```


```{r, echo = FALSE}
print(kable(head(apartments)))
```

### 2.2.1 Model 1: Linear regression

```{r, results = 'markup'}
apartments_lm_model <- lm(m2.price ~ construction.year + surface + floor +
                         no.rooms + district, data = apartments)
summary(apartments_lm_model)
```

```{r, echo = FALSE}
addHook(apartments_lm_model)
```


```{r}
predicted_mi2_lm <- predict(apartments_lm_model, apartmentsTest)
sqrt(mean((predicted_mi2_lm - apartmentsTest$m2.price)^2))
```


```{r}
explainer_lm <- explain(apartments_lm_model,
                          data = apartmentsTest[,2:6], y = apartmentsTest$m2.price)
```

```{r, echo = FALSE}
addHook(explainer_lm)
```

### 2.2.2 Model 2: Random forest

```{r, results = 'markup'}
library("randomForest")
set.seed(59)

apartments_rf_model <- randomForest(m2.price ~ construction.year + surface + floor +
                      no.rooms + district, data = apartments)
apartments_rf_model
```

```{r, echo = FALSE}
addHook(apartments_rf_model)
```


```{r}
predicted_mi2_rf <- predict(apartments_rf_model, apartmentsTest)
sqrt(mean((predicted_mi2_rf - apartmentsTest$m2.price)^2))
```

```{r}
explainer_rf <- explain(apartments_rf_model,
                       data = apartmentsTest[,2:6], y = apartmentsTest$m2.price)
```

```{r, echo = FALSE}
addHook(explainer_rf)
```

# Chapter 3 Model understanding

```{r}
library("DALEX")
apartments_lm_model <- lm(m2.price ~ construction.year + surface + floor +
                      no.rooms + district, data = apartments)

library("randomForest")
set.seed(59)
apartments_rf_model <- randomForest(m2.price ~ construction.year + surface + floor +
                      no.rooms + district, data = apartments)
```

```{r, echo = FALSE}
addHook(apartments_lm_model)
addHook(apartments_rf_model)
```

```{r}
explainer_lm <- explain(apartments_lm_model,
                       data = apartmentsTest[,2:6], y = apartmentsTest$m2.price)

explainer_rf <- explain(apartments_rf_model,
                       data = apartmentsTest[,2:6], y = apartmentsTest$m2.price)
```

```{r, echo = FALSE}
addHook(explainer_lm)
addHook(explainer_rf)
```

## 3.1 Model performance

```{r}
predicted_mi2_lm <- predict(apartments_lm_model, apartmentsTest)
sqrt(mean((predicted_mi2_lm - apartmentsTest$m2.price)^2))

predicted_mi2_rf <- predict(apartments_rf_model, apartmentsTest)
sqrt(mean((predicted_mi2_rf - apartmentsTest$m2.price)^2))
```


```{r, results='markup'}
mp_lm <- model_performance(explainer_lm)
mp_rf <- model_performance(explainer_rf)
mp_lm
mp_rf
```

```{r, echo = FALSE}
addHook(mp_lm)
addHook(mp_rf)
```


```{r}
plot(mp_lm, mp_rf)
```

```{r, echo = FALSE}
plot <- plot(mp_lm, mp_rf)
addHook(plot)
```

```{r}
plot(mp_lm, mp_rf, geom = "boxplot")
```

```{r, echo = FALSE}
plot <- plot(mp_lm, mp_rf, geom = "boxplot")
addHook(plot)
```

## 3.2 Feature importance

### 3.2.1 Model agnostic

```{r, results='markup'}
vi_rf <- variable_importance(explainer_rf, loss_function = loss_root_mean_square)
vi_rf
```

```{r, echo = FALSE}
addHook(vi_rf)
```

```{r, results='markup'}
vi_lm <- variable_importance(explainer_lm, loss_function = loss_root_mean_square)
vi_lm
```

```{r, echo = FALSE}
addHook(vi_lm)
```

```{r}
plot(vi_lm, vi_rf)
```

```{r, echo = FALSE}
plot <- plot(vi_lm, vi_rf)
addHook(plot)
```

```{r}
vi_lm <- variable_importance(explainer_lm, loss_function = loss_root_mean_square, type = "difference")
vi_rf <- variable_importance(explainer_rf, loss_function = loss_root_mean_square, type = "difference")
plot(vi_lm, vi_rf)
```

```{r, echo = FALSE}
addHook(vi_lm)
addHook(vi_rf)
plot <- plot(vi_lm, vi_rf)
addHook(plot)
```

### 3.2.2 Model specific

```{r, fig.show = FALSE, results = 'hide'}
varImpPlot(apartments_rf_model)
```

```{r, echo = FALSE}
plot <- varImpPlot(apartments_rf_model)
addHook(plot)
```

```{r}
library("forestmodel")
forest_model(apartments_lm_model)
```

```{r, echo = FALSE}
plot <- forest_model(apartments_lm_model)
addHook(plot)
```

```{r}
library("sjPlot")
plot_model(apartments_lm_model, type = "est", sort.est = TRUE)
```

```{r, echo = FALSE}
plot <- plot_model(apartments_lm_model, type = "est", sort.est = TRUE)
addHook(plot)
```


## 3.3 Variable response

### 3.3.1 Partial Dependence Plot


```{r}
sv_rf  <- single_variable(explainer_rf, variable =  "construction.year", type = "pdp")
plot(sv_rf)
```

```{r, echo = FALSE}
addHook(sv_rf) 
plot <- plot(sv_rf)
addHook(plot)
```

```{r}
sv_lm  <- single_variable(explainer_lm, variable =  "construction.year", type = "pdp")
plot(sv_rf, sv_lm)
```

```{r, echo = FALSE}
addHook(sv_lm)
plot <- plot(sv_rf, sv_lm)
addHook(plot)
```

### 3.3.2 Accumulated Local Effects Plot

```{r ale}
sva_rf  <- single_variable(explainer_rf, variable = "construction.year", type = "ale")
sva_lm  <- single_variable(explainer_lm, variable = "construction.year", type = "ale")

plot(sva_rf, sva_lm)
```

```{r, echo = FALSE}
addHook(sva_rf)
addHook(sva_lm)
plot <- plot(sva_rf, sva_lm)
addHook(plot)
```


### 3.3.3 Mering Path Plot


```{r factorMerger}
svd_rf  <- single_variable(explainer_rf, variable = "district", type = "factor")
svd_lm  <- single_variable(explainer_lm, variable = "district", type = "factor")

plot(svd_rf, svd_lm)
```

```{r, echo = FALSE}
addHook(svd_rf)
addHook(svd_lm)
plot <- plot(svd_rf, svd_lm)
addHook(plot)
```

# Chapter 4 Prediction understanding



```{r}
library("DALEX")
apartments_lm_model <- lm(m2.price ~ construction.year + surface + floor +
                      no.rooms + district, data = apartments)
library("randomForest")
set.seed(59)
apartments_rf_model <- randomForest(m2.price ~ construction.year + surface + floor +
                      no.rooms + district, data = apartments)
```

```{r, echo = FALSE}
addHook(apartments_lm_model)
addHook(apartments_rf_model)
```


```{r}
explainer_lm <- explain(apartments_lm_model,
                       data = apartmentsTest[,2:6], y = apartmentsTest$m2.price)
explainer_rf <- explain(apartments_rf_model,
                       data = apartmentsTest[,2:6], y = apartmentsTest$m2.price)
```

```{r, echo = FALSE}
addHook(explainer_lm)
addHook(explainer_rf)
```

## 4.1 Outlier detection

```{r}
mp_rf <- model_performance(explainer_rf)

library("ggplot2")
ggplot(mp_rf, aes(observed, diff)) + geom_point() +
        xlab("Observed") + ylab("Predicted - Observed") +
        ggtitle("Diagnostic plot for the random forest model") + theme_mi2()
```

```{r, echo = FALSE}
addHook(mp_rf)
plot <- ggplot(mp_rf, aes(observed, diff)) + geom_point() +
        xlab("Observed") + ylab("Predicted - Observed") +
        ggtitle("Diagnostic plot for the random forest model") + theme_mi2()
addHook(plot)
```

```{r, results='markup'}
which.min(mp_rf$diff)
## 1161
new_apartment <- apartmentsTest[which.min(mp_rf$diff), ]
new_apartment
```


```{r, echo = FALSE}
addHook(new_apartment)
```

4.2 Prediction breakDown

```{r, results='markup'}
new_apartment_rf <- single_prediction(explainer_rf, observation = new_apartment)
breakDown:::print.broken(new_apartment_rf)
```

```{r, echo = FALSE}
addHook(new_apartment_rf)
```


```{r}
plot(new_apartment_rf)
```

```{r, echo = FALSE}
plot <- plot(new_apartment_rf)
addHook(plot)
```

```{r}
new_apartment_lm <- single_prediction(explainer_lm, observation = new_apartment)
plot(new_apartment_lm, new_apartment_rf)
```

```{r, echo = FALSE}
addHook(new_apartment_lm)
plot <- plot(new_apartment_lm) 
addHook(plot)
```

# Chapter 5 Epilogue


```{r}
library("DALEX")

apartments_lm_model_improved <- lm(m2.price ~ I(construction.year < 1935 | construction.year > 1995) + surface + floor +
                         no.rooms + district, data = apartments)

explainer_lm_improved <- explain(apartments_lm_model_improved,
                          data = apartmentsTest[,2:6], y = apartmentsTest$m2.price)

mp_lm_improved <- model_performance(explainer_lm_improved)
plot(mp_lm_improved, geom = "boxplot")
```

```{r, echo = FALSE}
addHook(apartments_lm_model_improved)
addHook(explainer_lm_improved)
addHook(mp_lm_improved)
plot <- plot(mp_lm_improved, geom = "boxplot")
addHook(plot)
```

# Chapter 6 Exercises

```{r, results = 'hide'}
library("DALEX")
library("gbm")

apartments_gbm_model <- gbm(m2.price ~ construction.year + surface + floor +
                         no.rooms + district, data = apartments, n.trees = 1000)

explainer_gbm <- explain(apartments_gbm_model,
                          data = apartmentsTest[,2:6], y = apartmentsTest$m2.price,
                         predict_function = function(m, d) predict(m, d, n.trees = 1000))
```

```{r, echo = FALSE}
addHook(apartments_gbm_model)
addHook(explainer_gbm)
```

```{r, results = 'hide'}
library("nnet")
apartments_nnet_model <- nnet(m2.price ~ construction.year + surface + floor +
                         no.rooms + district, data = apartments,
                         linout=TRUE,
                         size = 50, maxit=100)

explainer_nnet <- explain(apartments_nnet_model,
                          data = apartmentsTest[,2:6], y = apartmentsTest$m2.price)
```

```{r, echo = FALSE}
addHook(apartments_nnet_model)
addHook(explainer_nnet)
```

```{r}
library("e1071")
apartments_svm_model <- svm(m2.price ~ construction.year + surface + floor +
                         no.rooms + district, data = apartments)

explainer_svm <- explain(apartments_svm_model,
                          data = apartmentsTest[,2:6], y = apartmentsTest$m2.price)

```


```{r, echo = FALSE}
addHook(apartments_svm_model)
addHook(explainer_svm)
```

```{r}
library("caret")
mapartments <- model.matrix(m2.price ~ ., data = apartments)
mapartmentsTest <- model.matrix(m2.price ~ ., data = apartmentsTest)
apartments_knn_model <- knnreg(mapartments, apartments[,1], k = 5)

explainer_knn <- explain(apartments_knn_model,
                          data = mapartmentsTest, y = apartmentsTest$m2.price)
```

```{r, echo = FALSE}
addHook(apartments_knn_model)
addHook(explainer_knn)
```



```{r}
mp_knn <- model_performance(explainer_knn)
mp_svm <- model_performance(explainer_svm)
mp_gbm <- model_performance(explainer_gbm)
mp_nnet <- model_performance(explainer_nnet)
plot(mp_gbm, mp_nnet, mp_svm, mp_knn, geom = "boxplot")
```


```{r, echo = FALSE}
addHook(mp_knn)
addHook(mp_svm)
addHook(mp_gbm)
addHook(mp_nnet)
plot <- plot(mp_gbm, mp_nnet, mp_svm, mp_knn, geom = "boxplot")
addHook(plot)
```




