# Conditional structure

## Partial Dependence Plots

Partial Dependence Plots (see **pdp** package [@pdp]) for a black box $f(x; \theta)$ calculates the expected output given a selected variable.

$$
p_i(x_i) = E_{x_{-i}}[ f(x^i, x^{-i}; \theta) ]
$$

Of course this expectation cannot be calculated directly as we do not know fully the $f()$ neither the distribution of $x_{-i}$. This value is estimated by 

$$
\hat p_i(x_i) = \frac{1}{n} \sum_{j=1}^{n} f(x^i_j, x_j^{-i}, \hat \theta) 
$$


Let's see an example for the model `HR_rf_model`. Below we are using `pdp::partial` function to calculate pdp curve for variable `satisfaction_level`. Then the curve is plotted with `plotPartial` (based on `lattice`) or `autoplot` (based on `ggplot2`).

```{r, message=FALSE, warning=FALSE}
library("pdp")
library("randomForest")
library("breakDown")

HR_rf_model <- randomForest(left~., data = breakDown::HR_data, ntree = 100)

part <- partial(HR_rf_model, "satisfaction_level")

plotPartial(part)

library("ggplot2")
autoplot(part)
```

## Individual Conditional Expectation Plot

Individual Conditional Expectations (ICE) may be considered as an extension of the PDP curves (see **ICEbox** package [@ICEbox]).
Instead of plotting expected value over all observations, for ICE we are plotting individual conditional model responses. Average of ICE curves results in PDP curve.

An ICE curve for observation $k$ over variable $i$ may be defined as

$$
ice_k(x_i) = f(x^i, x_k^{-i}; \theta) 
$$

ICE curves can be plotted with `pdp` package. Note that curves may be cantered in a given point, this helps in identification of possible interactions.

```{r, message=FALSE, warning=FALSE}
library("pdp")
library("randomForest")
library("breakDown")
library("ggplot2")

HR_rf_model <- randomForest(left~., data = breakDown::HR_data, ntree = 100)

part_rf_satisfaction <- partial(HR_rf_model, "satisfaction_level")

part_rf_satisfaction <- partial(HR_rf_model, pred.var = "satisfaction_level", ice = TRUE)
plotPartial(part_rf_satisfaction, rug = TRUE, train = HR_data, alpha = 0.2)
autoplot(part_rf_satisfaction, center = TRUE, alpha = 0.2, rug = TRUE, train = HR_data)
```

Or with the `ICEbox` package.

```{r, message=FALSE, warning=FALSE}
library("ICEbox")
part_rf_satisfaction = ice(object = HR_rf_model, X = HR_data, y = HR_data$satisfaction_level, predictor = "satisfaction_level", frac_to_build = .1)
plot(part_rf_satisfaction)
```

As ICE curves are useful tool for identification of interactions, these individual curves may be clustered with the `clusterICE` function.

```{r, message=FALSE, warning=FALSE}
clusterICE(part_rf_satisfaction, nClusters = 3, plot_legend = TRUE, center = TRUE)
```

## Factor Merger

The package `ICEbox` is not working for factor variables while the `pdp` package returns plots that are hard to interpret.

An interesting tool that helps to understand what is happening with factor variables is the **factorMerger** package (see [@factorMerger]).

Here we have Merging Path Plot for a factor variable `sales`.

```{r, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
library("factorMerger")
path <- mergeFactors(HR_data$left, HR_data$sales, method = "fast-adaptive", family = "binomial", abbreviate = FALSE)
plot(path, panel = "all")
```

Note that you can use the `factorMerger` package to understand predictions calculated with a black-box model. The random forest model `HR_rf_model` returns continuous response. But the `factorMerger` works for such variables as well.

In the top right panel one mey see the distribution of predictions for the selected group.

```{r, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}

HR_data$left_predicted <- predict(HR_rf_model)

path <- mergeFactors(HR_data$left_predicted, HR_data$sales, method = "fast-adaptive",  abbreviate = FALSE)
plot(path, panel = "all", responsePanel = "boxplot")
```

## ALEPlot

**ALEPlot** package [@ALEPlot]


