---
title: "How to use DALEX with the xgboost package"
author: 
  - Przemyslaw Biecek, Anna Kozak
date: First created on Nov 17, 2019. Updated on Nov 08, 2020
output: 
  html_document:
    toc: true  
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: kate
    fig_caption: true
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This vigniette demonstrates how to use the `DALEX` package with models created with the [xgboost](https://CRAN.R-project.org/package=xgboost) package. 

# Regression

In this example we are going to use the `wine` dataset from the `breakDown` package. The wine quality will be predicted based on other features.

```{r, warning=FALSE, message=FALSE}
library("breakDown")
head(wine)
```

## Model building

Let's build a model. We need to prepare `xgb.DMatrix` first.

```{r, warning=FALSE, message=FALSE}
library("xgboost")

model_martix_train <- model.matrix(quality ~ . - 1, wine)
data_train <- xgb.DMatrix(model_martix_train, label = wine$quality)
param <- list(max_depth = 2, eta = 1, silent = 1, nthread = 2,
              objective = "reg:linear")

wine_xgb_model <- xgb.train(param, data_train, nrounds = 50)
wine_xgb_model
```

## Explainer

Now we can create an explainer.

```{r, warning=FALSE, message=FALSE}
library("DALEX")

explainer_xgb <- explain(wine_xgb_model, 
                         data = model_martix_train, 
                         y = wine$quality, 
                         label = "xgboost",
                         colorize = FALSE)
explainer_xgb
```

## Single variable

For continouse variable

```{r, warning=FALSE, message=FALSE, fig.width=10, fig.height=8}
sv_xgb_satisfaction_level  <- model_profile(explainer_xgb, 
                                            variable = "alcohol", 
                                            type = "partial")

plot(sv_xgb_satisfaction_level)
```

## Single prediction

```{r, warning=FALSE, message=FALSE, fig.width=10, fig.height=6}
nobs <- model_martix_train[1, , drop = FALSE]
sp_xgb  <- predict_parts(explainer_xgb, 
                         new_observation = nobs,
                         type = "break_down")
head(sp_xgb)

plot(sp_xgb)
```

## Variable importance

```{r, warning=FALSE, message=FALSE, fig.width=10, fig.height=6}
vd_xgb <- model_parts(explainer_xgb, type = "raw")
head(vd_xgb)

plot(vd_xgb)
```


# Classification

In this example we are going to use the `HR_data` dataset from the `breakDown` package. The model will predict odds that someone will leave the company.

```{r, warning=FALSE, message=FALSE}
library("breakDown")
head(HR_data)
```

## Model building

Let's build a model. We need to prepare `xgb.DMatrix` first.

```{r, warning=FALSE, message=FALSE}
library("xgboost")

model_martix_train <- model.matrix(left ~ . - 1, HR_data)
data_train <- xgb.DMatrix(model_martix_train, label = HR_data$left)
param <- list(max_depth = 2, eta = 1, silent = 1, nthread = 2,
              objective = "binary:logistic", eval_metric = "auc")


HR_xgb_model <- xgb.train(param, data_train, nrounds = 50)
HR_xgb_model
```

## Explainer

Now we can create an explainer.

```{r, warning=FALSE, message=FALSE}
library("DALEX")
predict_logit <- function(model, x) {
  raw_x <- predict(model, x)
  exp(raw_x)/(1 + exp(raw_x))
}
logit <- function(x) exp(x)/(1+exp(x))

explainer_xgb <- explain(HR_xgb_model, 
                         data = model_martix_train, 
                         y = HR_data$left, 
                         predict_function = predict_logit,
                         link = logit,
                         label = "xgboost",
                         colorize = FALSE)
explainer_xgb
```

## Single variable

For continouse variable

```{r, warning=FALSE, message=FALSE, fig.width=10, fig.height=8}
sv_xgb_satisfaction_level  <- model_profile(explainer_xgb, 
                          variable = "satisfaction_level",
                          type = "partial")

plot(sv_xgb_satisfaction_level)
```

## Single prediction

```{r, warning=FALSE, message=FALSE, fig.width=10, fig.height=6}
nobs <- model_martix_train[1, , drop = FALSE]
sp_xgb  <- predict_parts(explainer_xgb, 
                         new_observation = nobs,
                         type = "break_down")
head(sp_xgb)

plot(sp_xgb)
```

## Variable importance

```{r, warning=FALSE, message=FALSE, fig.width=10, fig.height=6}
vd_xgb <- model_parts(explainer_xgb, type = "raw")
head(vd_xgb)

plot(vd_xgb)
```

# Session info

```{r, warning=FALSE, message=FALSE}
sessionInfo()
```

