---
title: "iBreak-down Plots for Instance Level Attributions with Interactions"
subtitle: "Code snippets for R "
author: "Przemysław Biecek"
date: "for DALEX 1.0"
output: 
  tint::tintHtml:
    toc: true
link-citations: yes
---

```{r setup, include=FALSE}
library(tint)
set.seed(1313)
```

In this section, we use an `DALEX::predict_parts()` function for calculation of Break Down plots with interactions.

If you want learn more about these plots read https://pbiecek.github.io/ema/iBreakDown.html.

# Prepare data

In this example we will use the titanic data.
It has few variables that are easy to understand.
The `titanic_imputed` dataset from `DALEX` package has imputed missing values.
Find more about the data https://pbiecek.github.io/ema/dataSetsIntro.html#TitanicDataset

```{r, warning=FALSE, message=FALSE}
library("DALEX")
head(titanic_imputed)
```

# Train a model

Here we use `ranger` library to train a classification random forest model.

```{r, warning=FALSE, message=FALSE}
library("ranger")
titanic_rf <- ranger(survived ~ class + gender + age + sibsp + 
         parch + fare + embarked, data = titanic_imputed,
         probability = TRUE,
         classification = TRUE)
titanic_rf
```

# Prepare an explainer

Different models have different structures.
We use `DALEX::explain()` function to create an uniform interface for the model 

```{r, warning=FALSE, message=FALSE}
library("DALEX")
titanic_ex <- explain(titanic_rf,
                data  = titanic_imputed,
                y     = titanic_imputed$survived,
                label = "Regression Forest for Titanic")
```

# Prepare an instance

For instance level explanations we need an observation for which we can generate explanation.

Let's create a data.frame with a single row that corresponds to 8 years old boy from 1st class.

```{r, warning=FALSE, message=FALSE}
henry <- data.frame(
         class = factor("2nd", levels = c("1st", "2nd", "3rd", "deck crew", 
                     "engineering crew", "restaurant staff", "victualling crew")),
         gender = factor("male", levels = c("female", "male")),
         age = 15,
         sibsp = 0,
         parch = 0,
         fare = 100,
         embarked = factor("Cherbourg", levels = c("Belfast",
                           "Cherbourg","Queenstown","Southampton"))
)
```

The predicted survival for `henry` is

```{r, warning=FALSE, message=FALSE}
predict(titanic_ex, henry)
```

# iBreak-down plots with `predict_parts()` 

The `DALEX::predict_parts()` function calculates the variable attributions for a selected model and the instance of interest. 

For BD with interactions this function requires three arguments: 

* the model explainer, 
* the data frame with the instance of interest, 
* the method for calculation of variable attribution, here `break_down_interactions` and

The resulting object is a data frame with variable contributions. 

```{r, warning=FALSE, message=FALSE}
library("DALEX")
ibd_rf <- predict_parts(titanic_ex,
                 new_observation = henry,
                 type = "break_down_interactions")
ibd_rf
```

Now we can plot this object with the generic `plot()` function. 

```{r, warning=FALSE, message=FALSE}
plot(ibd_rf) 
```

Let's see how this interaction was identified. Here we have Break down plots with two different orderings.

```{r, warning=FALSE, message=FALSE}
library("patchwork")
bd_rf_fag <- predict_parts(titanic_ex,
                 new_observation = henry,
                 order = c("class", "fare", "gender", "age", "embarked", "sibsp", "parch"))
bd_rf_afg <- predict_parts(titanic_ex,
                 new_observation = henry,
                 order = c("fare", "class", "gender", "age", "embarked", "sibsp", "parch"))
plot(bd_rf_afg) / plot(bd_rf_fag)
```
