---
title: "Shapley Additive Explanations (SHAP) for Instance Level Attributions"
subtitle: "Code snippets for R "
author: "Przemysław Biecek"
date: "for DALEX 1.0"
output: 
  tint::tintHtml:
    toc: true
link-citations: yes
---

```{r setup, include=FALSE}
library(tint)
set.seed(1313)
```

In this section, we use an `DALEX::predict_parts()` function for calculation of Shapley Values.
Note that there are also other R packages that offer similar functionality, like `shapper` which is a wrapper for the Python library `SHAP`, and `iml`.

If you want learn more about Shapley Values read https://pbiecek.github.io/ema/shapley.html.

# Prepare data

In this example we will use the titanic data.
It has few variables that are easy to understand.
The `titanic_imputed` dataset from `DALEX` package has imputed missing values.
Find more about the data https://pbiecek.github.io/ema/dataSetsIntro.html#TitanicDataset

```{r, warning=FALSE, message=FALSE}
library("DALEX")
head(titanic_imputed)
```

# Train a model

Here we use `ranger` library to train a classification random forest model.

```{r, warning=FALSE, message=FALSE}
library("ranger")
titanic_rf <- ranger(survived ~ class + gender + age + sibsp + 
         parch + fare + embarked, data = titanic_imputed,
         probability = TRUE,
         classification = TRUE)
titanic_rf
```

# Prepare an explainer

Different models have different structures.
We use `DALEX::explain()` function to create an uniform interface for the model 

```{r, warning=FALSE, message=FALSE}
library("DALEX")
titanic_ex <- explain(titanic_rf,
                data  = titanic_imputed,
                y     = titanic_imputed$survived,
                label = "Regression Forest for Titanic")
```

# Prepare an instance

For instance level explanations we need an observation for which we can generate explanation.

Let's create a data.frame with a single row that corresponds to 8 years old boy from 1st class.

```{r, warning=FALSE, message=FALSE}
johny_d <- data.frame(
            class = factor("1st", levels = c("1st", "2nd", "3rd", "deck crew",
                        "engineering crew", "restaurant staff", "victualling crew")),
            gender = factor("male", levels = c("female", "male")),
            age = 8,
            sibsp = 0,
            parch = 0,
            fare = 72,
            embarked = factor("Southampton", levels = c("Belfast",
                        "Cherbourg","Queenstown","Southampton")))
```

The predicted survival for `johny_d` is

```{r, warning=FALSE, message=FALSE}
predict(titanic_ex, johny_d)
```

# Shapley values with `predict_parts()` 

The `DALEX::predict_parts()` function calculates the variable attributions for a selected model and the instance of interest. 

For Shapley values this function requires four arguments: 

* the model explainer, 
* the data frame with the instance of interest, 
* the method for calculation of variable attribution, here `shap` and
* number of random orders that are used for Shap calculations. By default `B = 25`.

The resulting object is a data frame with variable contributions computed for every `B` ordering. Applying the generic function `plot()` to the object constructs the plot that includes the Shapley values and the corresponding box-plots.

```{r, warning=FALSE, message=FALSE}
shap_johny <- variable_attribution(titanic_ex, 
                                   johny_d, 
                                   type = "shap",
                                   B = 25)
shap_johny
```

To obtain a plot with only Shapley values, we can use the generic `plot()` function.

```{r, warning=FALSE, message=FALSE}
plot(shap_johny) 
```

To obtain a plot with only Shapley values, we can use the `show_boxplots=FALSE` argument in the `plot()` function call.

```{r, warning=FALSE, message=FALSE}
plot(shap_johny, show_boxplots = FALSE) 
```

It is the `ggplot2` object, so one can modify it as any other `ggplot2` object.

```{r, warning=FALSE, message=FALSE}
library("ggplot2")
plot(shap_johny, show_boxplots = FALSE) +
  ggtitle("Shapley values for Johny D") +
  theme(panel.grid = element_blank())
```

