---
title: "Ceteris-paribus Profiles for Instance Level Profiles"
subtitle: "Code snippets for R "
author: "Przemysław Biecek"
date: "for DALEX 1.0"
output: 
  tint::tintHtml:
    toc: true
link-citations: yes
---

```{r setup, include=FALSE}
library(tint)
set.seed(1313)
```

In this section, we use an `DALEX::predict_profile()` function for calculation of Ceteris-paribus profiles.

If you want learn more about Ceteris-paribus profiles read https://pbiecek.github.io/ema/ceterisParibus.html.

Note that there are also other R packages that offer similar functionality, like `condvis`, `pdp`, `ICEbox`, `ALEPlot` or `iml`.

# Prepare data

In this example we will use the titanic data.
It has few variables that are easy to understand.
The `titanic_imputed` dataset from `DALEX` package has imputed missing values.
Find more about the data https://pbiecek.github.io/ema/dataSetsIntro.html#TitanicDataset

```{r, warning=FALSE, message=FALSE}
library("DALEX")
head(titanic_imputed)
```

# Train a model

Here we use `ranger` library to train a classification random forest model.

```{r, warning=FALSE, message=FALSE}
library("ranger")
titanic_rf <- ranger(survived ~ class + gender + age + sibsp + 
         parch + fare + embarked, data = titanic_imputed,
         probability = TRUE,
         classification = TRUE)
titanic_rf
```

# Prepare an explainer

Different models have different structures.
We use `DALEX::explain()` function to create an uniform interface for the model 

```{r, warning=FALSE, message=FALSE}
library("DALEX")
titanic_ex <- explain(titanic_rf,
                data  = titanic_imputed,
                y     = titanic_imputed$survived,
                label = "Regression Forest for Titanic")
```

# Prepare an instance

For instance level explanations we need an observation for which we can generate explanation.

Let's create a data.frame with a single row that corresponds to 8 years old boy from 1st class.

```{r, warning=FALSE, message=FALSE}
johny_d <- data.frame(
            class = factor("1st", levels = c("1st", "2nd", "3rd", "deck crew",
                        "engineering crew", "restaurant staff", "victualling crew")),
            gender = factor("male", levels = c("female", "male")),
            age = 8,
            sibsp = 0,
            parch = 0,
            fare = 72,
            embarked = factor("Southampton", levels = c("Belfast",
                        "Cherbourg","Queenstown","Southampton")))
```

The predicted survival for `johny_d` is

```{r, warning=FALSE, message=FALSE}
predict(titanic_ex, johny_d)
```

# Ceteris-paribus profile with `predict_profile()` 

The `DALEX::predict_profile()` function calculates the conditional profiles for a selected model and the instance of interest. 

The resulting object is a data frame with CP profiles. Applying the generic function `plot()` to the object results in a profiles plotted with `ggplot2`.

The `DALEX::predict_profile()` function expects three arguments

* the model explainer, 
* the data frame with the instance of interest, 
* vector with names of variables for which CP profiles shall be calculated (by default all, but this may be time consuming).

 As a result, the function yields an object of the class `ceteris_paribus_explainer`. It is a data frame with model predictions.
 
```{r, warning=FALSE, message=FALSE}
cp_johny <- predict_profile(titanic_ex, 
                             johny_d)
cp_johny
```

To plot CP profiles, we can use the generic `plot()` function. The optional argument `variables` selects variables that are to be plotted.

```{r, warning=FALSE, message=FALSE}
plot(cp_johny, variables = c("age", "fare")) 
```

It is the `ggplot2` object, so one can modify it as any other `ggplot2` object.

```{r, warning=FALSE, message=FALSE}
library("ggplot2")
plot(cp_johny, variables = c("age", "fare")) +
  ggtitle("Ceteris Paribus for Johny D") +
  theme(panel.grid = element_blank()) +
  ylab("expected survival")
```


By default, all numerical variables are plotted. 
To plot CP profiles for categorical variables, we need to list them in the `variables` argument. 

```{r, warning=FALSE, message=FALSE}
plot(cp_johny, variables = c("class", "gender")) 
```


# Advanced use of the `predict_profile()` 

The `predict_profile()` is a very flexible function. To better understand how it can be used, we briefly review its arguments.

* `x`, `data`, `predict_function`, `label` - information about a model. If `x` is created with the `DALEX::explain` function, then other arguments are extracted from `x`,
* `new_observation` - instance (one or more), for which we want to calculate CP profiles. It should be a data frame with same variables as in the validation data.
* `variables` - names of explanatory variables, for which CP profiles are to be calculated. By default, the profiles will be constructed for all variables, which may be time consuming.
* `variable_splits` - a list of values for  which CP profiles are to be calculated. By default, these are all values for categorical variables. For continuous variables, uniformly-placed values are selected; one can specify the number of the values with the `grid_points` argument (the default is 101).

The argument `variable_splits` specifies the variables (`age` and `fare`) for which CP profiles are to be calculated, together with the list of values at which the profiles are to be evaluated. 

```{r, warning=FALSE, message=FALSE}
variable_splits = list(age = seq(0, 70, 0.1),
                       fare = seq(0, 100, 0.1))

titanic_cp <- predict_profile(titanic_ex, johny_d,
                  variable_splits = variable_splits)
```

(ref:titanicCeterisProfile01CDesc) Ceteris-paribus profiles for `class` and `embarked` variables and the `titanic_rf_v6` model. Blue dot stands for `henry`.

```{r titanicCeterisProfile01C, warning=FALSE}
plot(titanic_cp, variables = c("age", "fare"))
```


In the example below, we present the code to create CP profiles for two passengers, `henry` and `johny_d`. We use the  `scale_color_manual` function to add names of passengers to the plot, and to control colors and positions.

```{r, warning=FALSE, message=FALSE}
henry <- data.frame(
         class = factor("2nd", levels = c("1st", "2nd", "3rd", "deck crew", 
                     "engineering crew", "restaurant staff", "victualling crew")),
         gender = factor("male", levels = c("female", "male")),
         age = 15,
         sibsp = 0,
         parch = 0,
         fare = 100,
         embarked = factor("Cherbourg", levels = c("Belfast",
                           "Cherbourg","Queenstown","Southampton")))
         
cp_henry <- predict_profile(titanic_ex, rbind(johny_d, henry))

plot(cp_henry, color = "_ids_") + 
  scale_color_manual(name = "Passenger:", breaks = 1:2, 
            values = c("#4378bf", "#8bdcbe"), 
            labels = c("henry" , "johny_d"))
```


# Champion-challenger analysis with `predict_profile()`

One of the most interesting uses of the explainers is comparison of CP profiles for two or more of models.

To illustrate this possibility, first, we have go to construct profiles for the models. In our illustration, for the sake of clarity, we limit ourselves just to two models: the logistic regression and random forest models for the Titanic data. Moreover, we only consider the `age` and `fare` variables. We use `henry` as the instance, for which predictions are of interest.

```{r, warning=FALSE, message=FALSE}
library("rms")
titanic_lmr <- lrm(survived ~ gender + rcs(age) + class +
         sibsp + parch + fare + embarked, titanic_imputed)
titanic_ex_lmr <- explain(titanic_lmr, 
                        titanic_imputed,
                        label = "Logistic regression with splines",
                        verbose = FALSE)
```


```{r, warning=FALSE, message=FALSE}
titanic_cp_lmr <- predict_profile(titanic_ex_lmr, johny_d, 
                         variable_splits = variable_splits)
```

Subsequently, we construct the plot for both models.

```{r, warning=FALSE, message=FALSE}
plot(titanic_cp_lmr, titanic_cp, color = "_label_", 
     variables = c("age", "fare")) +
     ggtitle("Ceteris Paribus Profiles for johny_d","")
```


