---
title: "Model performance for Model Level Explanations"
subtitle: "Code snippets for R "
author: "Przemysław Biecek"
date: "for DALEX 1.0"
output: 
  tint::tintHtml:
    toc: true
link-citations: yes
---

```{r setup, include=FALSE}
library(tint)
set.seed(1313)
```

In this section, we use an `DALEX::model_performance()` function for calculation of various model performance measures.

If you want learn more about mathematical formations of these measures read https://pbiecek.github.io/ema/modelPerformance.html.

Note that there are also other R packages that offer similar functionality, like `mlr`.

# Classification 

# Prepare data

For classification example we will use the titanic data.
It has few variables that are easy to understand.
The `titanic_imputed` dataset from `DALEX` package has imputed missing values.
Find more about the data https://pbiecek.github.io/ema/dataSetsIntro.html#TitanicDataset

```{r, warning=FALSE, message=FALSE}
library("DALEX")
head(titanic_imputed)
```

# Train models

Here we use `ranger` library to train a classification random forest model and `rms` model to train a logistic regression model with splines.

```{r, warning=FALSE, message=FALSE}
library("ranger")
titanic_rf <- ranger(survived ~ class + gender + age + sibsp + 
         parch + fare + embarked, data = titanic_imputed,
         probability = TRUE,
         classification = TRUE)
titanic_rf

library("rms")
titanic_lmr <- lrm(survived ~ gender + rcs(age) + class +
         sibsp + parch + fare + embarked, titanic_imputed)
```

# Prepare an explainer

Different models have different structures.
We use `DALEX::explain()` function to create an uniform interface for the model 

```{r, warning=FALSE, message=FALSE}
library("DALEX")
titanic_ex <- explain(titanic_rf,
                data  = titanic_imputed,
                y     = titanic_imputed$survived,
                label = "Regression Forest for Titanic",
                verbose = FALSE)

titanic_ex_lmr <- explain(titanic_lmr, 
                data  = titanic_imputed,
                y     = titanic_imputed$survived,
                label = "Logistic regression with splines",
                verbose = FALSE)
titanic_ex_lmr$model_info$type = "classification"
```

# Model performance for classification models with `model_performance()` 

The `DALEX::model_performance()` function calculates the most popular model measures.  This function expects two arguments

* the model explainer, 
* `cutoff` with the cutoff for measures that work on binary classifications rather than scores.

 As a result, the function yields an object of the class `model_performance`. It is a data frame with model performance measures

* recall, also known as sensitivity,
* precision, also called positive predictive value,
* f1 score is the harmonic mean of precision and recall,
* accuracy is the fraction of correct responses,
* auc area under the curve.

```{r, message=FALSE, echo=FALSE}
(titanic_mp     <- model_performance(titanic_ex))
(titanic_mp_lmr <- model_performance(titanic_ex_lmr))
```

Model performance measure may be supplemented with graphical summary, like ROC curve.

```{r, warning=FALSE, message=FALSE}
plot(titanic_mp, titanic_mp_lmr, geom = "roc")
```

Or the lift curve.

```{r, warning=FALSE, message=FALSE}
plot(titanic_mp, titanic_mp_lmr, geom = "lift")
```

Or the gain curve.

```{r, warning=FALSE, message=FALSE}
plot(titanic_mp, titanic_mp_lmr, geom = "gain")
```

Or boxplot

```{r, warning=FALSE, message=FALSE}
plot(titanic_mp, titanic_mp_lmr, geom = "boxplot")
```

Or histogram

```{r, warning=FALSE, message=FALSE}
plot(titanic_mp, titanic_mp_lmr, geom = "histogram")
```



# Regression 

# Prepare data

For regression  example we will use the `apartments` data from `DALEX` package.
It has few variables that are easy to understand.
Find more about the data https://pbiecek.github.io/ema/dataSetsIntro.html#ApartmentDataset

```{r, warning=FALSE, message=FALSE}
library("DALEX")
head(apartments)
```

# Train models

We will use two models in this example. A simple linear model trained with `lm()` function and random forest model trained with `randomForest()` function.

```{r, warning=FALSE, message=FALSE}
set.seed(13)

apartments_lm <- lm(m2.price ~ ., data = apartments)
anova(apartments_lm)

library("randomForest")
apartments_rf <- randomForest(m2.price ~ ., data = apartments)
apartments_rf
```

# Prepare explainers

Different models have different structures.
We use `DALEX::explain()` function to create an uniform interface for the model 

```{r, warning=FALSE, message=FALSE}
library("DALEX")
explain_lm <- explain(model = apartments_lm, 
         data = apartments_test, 
         y = apartments_test$m2.price,
         label = "Linear Regression for apartments",
         verbose = FALSE)

explain_rf <- explain(model = apartments_rf, 
         data = apartments_test, 
         y = apartments_test$m2.price,
         label = "Random Forest for apartments",
         verbose = FALSE)
```

# Model performance for regression models with `model_performance()` 

The `DALEX::model_performance()` function calculates the most popular model measures.  The only expected argument is the model explainer. As a result, the function yields an object of the class `model_performance`. 

It is a data frame with selected model performance measures

* mean square error,
* root mean square error,
* median absolute deviation,
* R-square coefficient.
 
```{r, message=FALSE, echo=FALSE}
(apartments_mp    <- model_performance(explain_lm))
(apartments_mp_rf <- model_performance(explain_rf))
```

Both plots can be supplemented with boxplots for residuals. Toward this end, the residuals have got to be computed and added to the explainer object with the help of the `model_performance()` function. Subsequently, the `plot()` can be applied to the resulting object.   

```{r, warning=FALSE, message=FALSE}
plot(apartments_mp, apartments_mp_rf, geom = "boxplot")
```

And the histogram.

```{r, warning=FALSE, message=FALSE}
plot(apartments_mp, apartments_mp_rf, geom = "histogram")
```

